A Parallel General Purpose Multi-Objective Optimization Framework, with Application To Beam Dynamics

### 2018-05-20 paper/paper.tex
editing awa section 5.3

The next section introduces the notation of multi-objective optimization
  theory and describes the first implemented optimizer.
In Section~\ref{sec:framework} we discuss the implementation of the framework.
We introduce the employed forward-solver in Section~\ref{sec:forward-solver}.
A validation and a proof of concept application in two beam dynamics problems is
  discussed in Section~\ref{sec:experiments}.

@@

The authors thank the \textsc{SwissFEL} and AWA teams for contributing to the
  formulation of optimization problems. 
  We gratefully acknowledge the computing resources provided on Bebop,
  a high performance computing cluster operated by the 
  Labratory Computing Resource Center at Argonne National Laboratory.
  This work was partly supported by the 
  U.S. Department of Energy, Office of Science, under 
  contract number DE-AC02-06CH11357 and grant number DE-SC0015479. 
\appendix

@@

\bibliographystyle{abbrv}
\bibliography{../../bibliography/phd}

>>>>

The next section introduces the notation of multi-objective optimization
  theory and describes the first implemented optimizer.
In Section~\ref{sec:framework} we discuss the implementation of the framework.
We introduce the employed forward-solver in Section~\ref{sec:forward-solver}.
A validation and a proof of concept application in the two beam dynamics problems 
mentioned above is discussed in Section~\ref{sec:experiments}.

@@

The authors thank the \textsc{SwissFEL} and AWA teams for contributing to the
  formulation of optimization problems. 
  We gratefully acknowledge the computing resources provided on Bebop,
  a HPC cluster operated by the LCRC at ANL.
  This work was partly supported by the 
  U.S. Department of Energy, Office of Science, under 
  contract number DE-AC02-06CH11357 and grant number DE-SC0015479. 
\appendix

@@

\bibliographystyle{abbrv}
\bibliography{paper}

### 2018-05-20 paper/theor-experiment.tex
editing awa section 5.3

%Currently we run the simulations on a low resolution space charge grid ($8
  %\times 8 \times 8$) as a proof of concept.
%In a next step a simple Python script would be applied to iteratively rerun
  %the optimizer with increasing space-charge resolution when the limits of the
  %design variables are narrow enough (iteratively diminishing search space). 
%\vspace{-7em}
\subsection{AWA Photoinjector Optimization} \label{awaproblem}
Next we apply the optimization framework to the high charge beam line at the AWA. 
As shown in Fig.  \ref{awa-pic}, this beam line consists of an rf photocathode gun, 
two solenoids, six linear accelerating cavities, 
followed by four quadrupoles and a stripline kicker. 
This beam line has served as the injector for Two Beam 
Acceleration (TBA) experiments in the past [X1], and will be used for 
similar experiments in the future [X2]. Main constraints on prior experimental 
results were the beam size and emittance as the beam passed through small aperture 
wakefield structures located after the quadrupoles. 
In an attempt to maximize charge transmission in upcoming experiments
we optimize several parameters in the gun and quadrupole leading into the 
TBA section of the beam line.
%\vspace{-1em}

@@

%\vspace{-1em}
We also chose this model to demonstrate the frameworks ability to tackle larger problems.
The number of design variables is eight, and the number of objectives is seven.
This problem is also less predictable, due to the high dimensionality 
and nonlinear effects such as space charge. There is not a definite idea
before hand where the Pareto front will fall. This work is also meaningful in that
it will guide future operations at the AWA.

@@

The first four objectives minimize the transverse ($rms_x$) and longitudinal ($rms_s$) 
beam sizes at two locations of interest in the beam line ($s_1$,$s_2$). 
These are the longitudinal positions at the entrance of the kicker and 
$1[\text{m}]$ after the exit of the kicker. The beam size needs to be minimized
here to prevent scraping and ensure transmission through the wakefield structures 
following the kicker. The next two objectives (\ref{eq:awa:p4}) minimize the 
transverse emittance, which favors less divergence in the beam and reduces 
emittance growth during transport. The last objective (\ref{eq:awa:p4}), 
minimizes energy spread in the bunch at the entrance of the kicker, $s_1$.  
This helps reduce emittance and beam size growth in the kicker.  
Equations (\ref{eq:awa:firstconstr}) to
(\ref{eq:awa:lastdvar}) define the charge, gun voltage, linac voltages, 
laser radius and linac cavity phases. These are parameters in the simulation 
that must be defined, but do not vary during the optimization.

@@

Design variables include the currents in two gun solenoids (IBF and IM), 
the gun phase (GPHASE), the laser longitudinal full width half max (FWHM), 
and four quadrupole strengths (KQ1-KQ2). The objectives include
beam size, emittance, and energy spread as
defined in Eqs. (\ref{eq:awa:p1}) to (\ref{eq:awa:p4}). 
The s location at the entrance of the kicker is $s_1=16.45[\text{m}]$, 
and the s location of interest after the kicker is $s_1=18.45[\text{m}]$.

All simulations for this experiment were carried out on Bebop a
high performance computing (HPC)
cluster provided by the Laboratory Computing Resource Center (LCRC)
at Argonne National Laboratory (ANL). We used Intel Knights Landing 
(KNL) processors at 1.3 GHz with 128 GB of memory 
and 64 cores per node. There are 352 compute nodes available on 
Bebop, with a total of 22,528 cores. All jobs were run and compared 
on 8 cores each, which allowed 8 jobs per node on the KNL.
This in combination with the amount of nodes available 
allows for very large optimization jobs, like the AWA case.

A study on time step and 
number of particles was done to reduce the time to simulation while 
maintaining the physics of interest. 
The grid size was $16 \times 16 \times 32$, 
and parallelized in the x and y directions.
After comparing several options we choose ten thousand particles 
in combination with adjusting the time step near sensitive elements 
(i.e. quadurpoles, kicker). 
The resulting simulations are low fidelity, but closely approximate 
the mid fidelity simulations for metrics of interest, as shown in 
Fig. \ref{tstep}. For all models, the longitudinal parameters (rms$_s$ and energy) 
are calculated correctly, but we see discrepancies in the transverse 
rms$_x$ and $\epsilon_x$ in the low fidelity results. A summary of 
the parameters used is listed in Table \ref{fidelity}.

\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{Report/timestep_comparison}
	\caption{Comparison of different fidelity models.}
	\label{tstep}
\end{figure}

\begin{table}%[h!]
	\begin{center}
		\caption{Low and mid fidelity simulation parameters.}
		\label{fidelity}
		\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} C c D }
			\hline\noalign{\smallskip}
			& Number of particles & dT (seconds) & Time to simulation (minutes)\\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			low fidelity  			&  10,000   & $5 \times10^{-11}$  &  2.6 \\
			adjusted low fidelity  	&  10,000   & $2 \times 10^{-12}$, $1 \times 10^{-11}$ & 1.6 \\
			mid fidelity 			&  100,000  & $1 \times 10^{-12}$ &  18\\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}

\subsubsection{AWA Optimization Results}
While the optimization problem and goals are well defined (Section \ref{awaproblem}), 
it is not clear what the best input parameters for the NSGA-II are.
These parameters include gene mutation probability, mutation probability, 
recombination probability, number of individuals (initial and and for later generations), and number of generations to complete. 
Given the beam line and design variables given in Section \ref{awaproblem}, 
four optimization experiments were done with varying NSGA-II parameters. 
Similar to the time step scan, 
the goal of this exercise was to determine which set of optimization
parameters strongly influence the results, 
and whether there was a time to solution difference.
From here on, we will reference each experiment as ex-1, ex-2, ex-3, and ex-4
as shown in Table \ref{extable}. 
\begin{table}%[h!]
	\begin{center}
		\caption{Input Parameters for initial twenty four hour AWA optimization experiments. 
			The gene mutation probability was equal to the mutation probability (not shown) in all four experiments. The number of individuals}
		\label{extable}
		\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} C C D }
			\hline\noalign{\smallskip}
			& Gene Mutation Probability & Recombination Probibility & Number of completed generations \\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			ex-1 &  0.1  & 0.9  &  96 \\
			ex-2 &  0.3  & 0.7  &  81 \\
			ex-3 &  0.8  & 0.2  &  53 \\
			ex-4 &  0.01 & 0.09 &  95 \\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}

The initial number of individuals was fixed at 656, 
and the minimum number individuals in later generations was fixed at 328. 
These numbers we in part based on the architecture of the KNL's. 
Since each simulation takes 8 cores, and there are 64 cores per KNL node, 
we wanted a large population size that would fit evenly on these resources. 
Each experiment was allowed to run for twenty four hours, with 
a max generation limit of 100. 
We also reduced the six objectives to the four most important for quicker evaluation of these tests \nrnote{Q for Andreas, is it true less obj is faster}.
The objectives include: 
$\varepsilon_{x}\left(s = s_1\right)\text{, } \varepsilon_{x}\left(s = s_2\right)\text{, } \text{rms}_{s}\left(s = s_1\right)\text{, and }  \text{rms}_{s}\left(s = s_2\right)$. 
The OPAL input file for ex-3 is given as an example:

>>>>

%Currently we run the simulations on a low resolution space charge grid ($8
  %\times 8 \times 8$) as a proof of concept.
%In a next step a simple Python script would be applied to iteratively rerun
  %the optimizer with increasing space-charge resolution when the limits of the
  %design variables are narrow enough (iteratively diminishing search space). 
%\vspace{-7em}
\subsection{AWA Photoinjector Optimization} \label{awaproblem}
Next we apply the optimization framework to the high charge beam line at the AWA. 
As shown in Fig.  \ref{awa-pic}, this beam line consists of an rf photocathode gun, 
two solenoids, six linear accelerating cavities, 
followed by four quadrupoles and a stripline kicker. 
The charge of interest is 40 nC, and based on Two Beam 
Acceleration (TBA) \cite{gai_power_jing_2012} experiments performed in the past 
at AWA. Similar experiments will take place in the future and motivate this work. 
Prior experimental results were limited by beam size and emittance as the beam passed through small aperture 
wakefield structures located after the quadrupoles. 
In an attempt to maximize charge transmission in upcoming experiments
we optimize several parameters in the gun and quadrupoles leading into the 
TBA section of the beam line.
%\vspace{-1em}

@@

%\vspace{-1em}
In addition to the physics challenges, we chose this model to demonstrate the frameworks ability to tackle large problems.
The number of design variables is eight, and the number of objectives is seven.
This problem encompasses high dimensionality 
and nonlinear effects such as space charge. There is no existing information
that accurately predicts where the Pareto front will fall. This work is also meaningful in that
it will guide future operations at the AWA.

@@

The first four objectives, Eqs. (\ref{eq:awa:p1}) and (\ref{eq:awa:p2}),
minimize the transverse ($rms_x$) and longitudinal ($rms_s$) 
beam sizes at two locations of interest in the beam line ($s_1$,$s_2$);
the longitudinal positions at the entrance of the kicker and 
$1$ meter after the exit of the kicker. 
Minimizing the beam size at these locations is essential to 
to preventing scraping and ensuring transmission through the wakefield structures 
following the kicker. The next two objectives (\ref{eq:awa:p3}) minimize the 
transverse emittance, which favors less divergence in the beam and reduces 
emittance growth during transport. The last objective (\ref{eq:awa:p4}), 
minimizes energy spread in the bunch at the entrance of the kicker at $s_1$.  
This helps reduce emittance and beam size growth in the kicker.  
Equations (\ref{eq:awa:firstconstr}) to
(\ref{eq:awa:lastdvar}) define the charge, gun voltage, linac voltages, 
laser radius and linac cavity phases (in that order). These are parameters in the simulation 
that must be defined, but do not vary during the optimization.

@@

Design variables include the currents in two gun solenoids (IBF and IM), 
the gun phase (GPHASE), the laser longitudinal full width half max (FWHM), 
and four quadrupole strengths (KQ1-KQ2). The objectives include
beam size, emittance, and energy spread as
defined in Eqs. (\ref{eq:awa:p1}) to (\ref{eq:awa:p4}). 
The location at the entrance of the kicker is $s_1=16.45$~meters, 
and the location of interest after the kicker is $s_1=18.45$~meters.

All simulations for this experiment were carried out on Bebop a
high performance computing (HPC)
cluster provided by the Laboratory Computing Resource Center (LCRC)
at Argonne National Laboratory (ANL). We used Intel Knights Landing 
(KNL) processors at 1.3 GHz with 128 GB of memory 
and 64 cores per node. There are 352 compute nodes available on 
Bebop, with a total of 22,528 cores. All jobs were run and compared 
on 8 cores each, which allowed 8 jobs per node on the KNLs.
This in combination with the amount of nodes available 
allows for very large optimization jobs, like the AWA case.

A study on time step and 
number of particles was done to reduce the time to simulation while 
maintaining the physics of interest. 
The grid size was $16 \times 16 \times 32$, 
and parallelized in the x and y directions.
After comparing several options we choose ten thousand particles 
in combination with adjusting the time step (dT) near sensitive elements 
(i.e. quadurpoles, kicker). 
The resulting simulations are low fidelity, but closely approximate 
the mid fidelity simulations for metrics of interest, as shown in 
Fig. \ref{tstep}. For all models, the longitudinal parameters (rms$_s$ and energy) 
are calculated correctly, but we see discrepancies in the transverse 
rms$_x$ and $\epsilon_x$ in the low fidelity results. 
Average run time each simulation with the adjusted time steps was 1.6~minutes.
In comparison the mid fidelity simulations ran for 18~minutes.
%A summary of the parameters used is listed in Table \ref{fidelity}.

\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{Report/timestep_comparison}
	\caption{Comparison of different fidelity models (dT stands for time step).}
	\label{tstep}
\end{figure}

\iffalse
\begin{table}%[h!]
	\begin{center}
		\caption{Low and mid fidelity simulation parameters.}
		\label{fidelity}
		\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} C c D }
			\hline\noalign{\smallskip}
			& Number of particles & dT (seconds) & Time to simulation (minutes)\\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			low fidelity  			&  10,000   & $5 \times10^{-11}$  &  2.6 \\
			adjusted low fidelity  	&  10,000   & $2 \times 10^{-12}$, $1 \times 10^{-11}$ & 1.6 \\
			mid fidelity 			&  100,000  & $1 \times 10^{-12}$ &  18\\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}
\fi

\subsubsection{AWA Optimization Results}
While the optimization problem and goals were well defined (Section \ref{awaproblem}), 
it was not clear what the best input parameters for the NSGA-II should be.
These parameters include gene mutation probability, mutation probability, 
recombination probability, number of individuals (initial and and for later generations), and number of generations to complete. 
Given the beam line and design variables given in Section \ref{awaproblem}, 
four optimization experiments were done with varying NSGA-II parameters. 
Similar to the time step scan, 
the goal of this exercise was to determine which set of optimization
parameters strongly influence the results, 
and whether there was a time to solution difference.
From here on, we will reference each experiment as ex-1, ex-2, ex-3, and ex-4
as shown in Table \ref{extable}. 
\begin{table}%[h!]
	\begin{center}
		\caption{Input Parameters for initial twenty four hour AWA optimization experiments. 
			The gene mutation probability was equal to the mutation probability (not shown) in all four experiments. 
			The max number of individuals per generation was~80.}
		\label{extable}
		\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} C C D }
			\hline\noalign{\smallskip}
			& Gene Mutation Probability & Recombination Probibility & Number of completed generations \\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			ex-1 &  0.1  & 0.9  &  96 \\
			ex-2 &  0.3  & 0.7  &  81 \\
			ex-3 &  0.8  & 0.2  &  53 \\
			ex-4 &  0.01 & 0.09 &  95 \\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}

The initial number of individuals was fixed at 656, 
and the minimum number individuals in later generations was fixed at 328. 
These numbers we in part based on the architecture of the KNL's. 
Since each simulation takes 8 cores, and there are 64 cores per KNL node, 
we wanted a large population size that would fit evenly on these resources. 
Each experiment was allowed to run for twenty four hours, with 
a max generation limit of 100. 
We also reduced the six objectives to the four most important for quicker evaluation of these tests 
%\nrnote{Q for Andreas, is it true less obj is faster}.
The objectives include: 
$\varepsilon_{x}\left(s = s_1\right)\text{, } \varepsilon_{x}\left(s = s_2\right)\text{, } \text{rms}_{s}\left(s = s_1\right)\text{, and }  \text{rms}_{s}\left(s = s_2\right)$. 
The OPAL input file for ex-3 is given as an example:
